Short-term hourly forecasting for care services

### 2021-10-04 paper/paper.Rmd
added admission viz

journal: "Which journal? Journal of Service Research/Health Services Research/EJOR/IJF"
date: "`r Sys.Date()`"
geometry: "top=25mm, left=30mm, right=30mm, bottom=25mm,headsep=10mm, footskip=12mm"
#linenumbers: true
numbersections: true
#output: rticles::elsevier_article
header-includes:  
 \usepackage{adjustbox, float,lscape} #use the 'float' package
output:
  bookdown::pdf_book: 
    base_format: rticles::elsevier_article
    citation_package: natbib
bibliography: mybibfile.bib
biblio-style: authoryear
#csl: elsevier-without-titles.csl
#biblio-style: abbrevnat
---

@@

If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters. Generalised additive models for location, scale and shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables. This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found. In practice, this means experimenting with various distributions and evaluating their suitability using available training data.

Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution. In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters. In a GAMLSS framework \cite{Rigby2005} the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via
\begin{equation} \label{eq:basicGAM}
    g_j(\theta_{j,t})=\mathbf{A}_{j,t} \bm{\beta}_j + \sum_{i} f_{j,i}({\bm x}^{S_{j,i}}_t), \;\;\; \text{for} \;\;\; j = 1, \dots, m,
\end{equation}
where $g_j$ is a monotonic `link' function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$. If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third element of $\bm x_t$. Each $f_{j,i}$ is a smooth function, constructed as
\begin{equation}
    f_{j,i}(\bm x^{S_{j,i}}) = \sum_{k=1}^{K_{j,i}} b^{ji}_k (\bm x^{S_{j,i}}) \beta_k^{ji},
\end{equation}
where $b^{ji}_k$ are spline basis functions of dimension $\vert S_{j,i} \vert$, while $\beta_k^{ji}$ are regression coefficients. The smoothness of each $f_{j,i}$ is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used. See \cite{Wood2017} for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.

As our data are counts, the natural starting point is the Poisson distribution, given by
\begin{equation}
  F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor}
\end{equation}
where we consider an additive model for $\lambda_t$ of the form
\begin{equation}
  \log(\lambda_t) = \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad .
\end{equation}
The functions $H(t)$, $D(t)$ and $Y(t)$ return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time $t$, respectively, and $C_t$ is the temperature at time $t$.

Truncated Normal...
\begin{equation}
  F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
\end{equation}
with
\begin{eqnarray}
  \mu_t &=& \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad , \\
  log(\sigma_t) &=& f(H(t))  \quad .
\begin{end}

Truncated $t$ distribution...

Negative binomial...

>>>>

journal: "Which journal? Journal of Service Research/Health Services Research/EJOR/IJF"
date: "`r Sys.Date()`"
geometry: "top=25mm, left=30mm, right=30mm, bottom=25mm,headsep=10mm, footskip=12mm"
#linenumbers: true
numbersections: true
#output: rticles::elsevier_article
header-includes:  
 \usepackage{adjustbox, float,lscape, bm} #use the 'float' package
output:
  bookdown::pdf_book: 
    base_format: rticles::elsevier_article
    citation_package: natbib
bibliography: mybibfile.bib
#biblio-style: authoryear
#csl: elsevier-without-titles.csl
#biblio-style: abbrevnat
---

@@

Let $y_t$ be the number of attendances in time period $t$ and indicate with $F_t(y_t)$ its predictive cumulative probability distribution. In a distributional regression context, $F_t(y_t)$ is modelled via a parametric model, $F(y_t|\bm \theta_t)$, where $\bm \theta_t$ is an $m$-dimensional vector of parameters. In a GAMLSS framework @Rigby2005 the elements $j=1,...,m$ of $\bm \theta_t$ are modelled via
\begin{equation} \label{eq:basicGAM}
    g_j(\theta_{j,t})=\mathbf{A}_{j,t} \bm{\beta}_j + \sum_{i} f_{j,i}({\bm x}^{S_{j,i}}_t), \;\;\; \text{for} \;\;\; j = 1, \dots, m,
\end{equation}
where $g_j$ is a monotonic function, $\mathbf{A}_{j,t}$ is the $t$-th row of the design matrix $\mathbf{A}_j$, $\bm \beta_j$ is a vector of regression coefficients, $\bm x_t$ is a $d$-dimensional vector of covariates and $S_{j,i} \subset \{1, \dots, d\}$. If $S_{j,i} = \{1, 3\}$, then following our notation ${\bm x}_{t}^{S_{j,i}}$ is a two dimensional vector formed by the first and third element of $\bm x_t$. Each $f_{j,i}$ is a smooth function, constructed as
\begin{equation}
    f_{j,i}(\bm x^{S_{j,i}}) = \sum_{k=1}^{K_{j,i}} b^{ji}_k (\bm x^{S_{j,i}}) \beta_k^{ji},
\end{equation}
where $b^{ji}_k$ are spline basis functions of dimension $\vert S_{j,i} \vert$, while $\beta_k^{ji}$ are regression coefficients. The smoothness of each $f_{j,i}$ is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used. See @Wood2017 for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.

### 2021-10-04 paper/paper.tex
added admission viz

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[top=25mm, left=30mm, right=30mm, bottom=25mm,headsep=10mm, footskip=12mm]{geometry}
\usepackage{natbib}
\bibliographystyle{authoryear}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Short-term hourly forecasting for care services},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header
\usepackage{adjustbox, float,lscape}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

If we assume that our predictive distribution follows a given parametric distribution, the forecasting task becomes ones of predicting the future values of that distribution's parameters. Generalised additive models for location, scale and shape (GAMLSS) are distributional regression models where the parameters of the assumed distribution may be modelled as additive functions of explanatory variables. This provides a powerful and flexible framework for probabilistic forecasting, provided that a suitable distribution and additive structures can be found. In practice, this means experimenting with various distributions and evaluating their suitability using available training data.

Let \(y_t\) be the number of attendances in time period \(t\) and indicate with \(F_t(y_t)\) its predictive cumulative probability distribution. In a distributional regression context, \(F_t(y_t)\) is modelled via a parametric model, \(F(y_t|\bm \theta_t)\), where \(\bm \theta_t\) is an \(m\)-dimensional vector of parameters. In a GAMLSS framework \cite{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled via
\begin{equation} \label{eq:basicGAM}
    g_j(\theta_{j,t})=\mathbf{A}_{j,t} \bm{\beta}_j + \sum_{i} f_{j,i}({\bm x}^{S_{j,i}}_t), \;\;\; \text{for} \;\;\; j = 1, \dots, m,
\end{equation}
where \(g_j\) is a monotonic `link' function, \(\mathbf{A}_{j,t}\) is the \(t\)-th row of the design matrix \(\mathbf{A}_j\), \(\bm \beta_j\) is a vector of regression coefficients, \(\bm x_t\) is a \(d\)-dimensional vector of covariates and \(S_{j,i} \subset \{1, \dots, d\}\). If \(S_{j,i} = \{1, 3\}\), then following our notation \({\bm x}_{t}^{S_{j,i}}\) is a two dimensional vector formed by the first and third element of \(\bm x_t\). Each \(f_{j,i}\) is a smooth function, constructed as
\begin{equation}
    f_{j,i}(\bm x^{S_{j,i}}) = \sum_{k=1}^{K_{j,i}} b^{ji}_k (\bm x^{S_{j,i}}) \beta_k^{ji},
\end{equation}
where \(b^{ji}_k\) are spline basis functions of dimension \(\vert S_{j,i} \vert\), while \(\beta_k^{ji}\) are regression coefficients. The smoothness of each \(f_{j,i}\) is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used. See \cite{Wood2017} for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.

As our data are counts, the natural starting point is the Poisson distribution, given by
\begin{equation}
  F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor}
\end{equation}
where we consider an additive model for \(\lambda_t\) of the form
\begin{equation}
  \log(\lambda_t) = \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad .
\end{equation}
The functions \(H(t)\), \(D(t)\) and \(Y(t)\) return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time \(t\), respectively, and \(C_t\) is the temperature at time \(t\).

Truncated Normal\ldots{}
\begin{equation}
  F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
\end{equation}
with
\textbackslash begin\{eqnarray\}
\mu\emph{t \&=\& \sum}\{i=1\}\^{}7 \beta\emph{i \delta(D\_i(t)-i) + \sum}\{j=1\}\^{}7 D\_j(t) f\_j(H(t)) + t f\_\text{Y}(Y(t)) + f\_\text{Temp}(Y(t),C\_t) \quad , \textbackslash{}
log(\sigma\_t) \&=\& f(H(t)) \quad .
\textbackslash begin\{end\}

Truncated \(t\) distribution\ldots{}

Negative binomial\ldots{}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[top=25mm, left=30mm, right=30mm, bottom=25mm,headsep=10mm, footskip=12mm]{geometry}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Short-term hourly forecasting for care services},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header
\usepackage{adjustbox, float,lscape, bm}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

Let \(y_t\) be the number of attendances in time period \(t\) and indicate with \(F_t(y_t)\) its predictive cumulative probability distribution. In a distributional regression context, \(F_t(y_t)\) is modelled via a parametric model, \(F(y_t|\bm \theta_t)\), where \(\bm \theta_t\) is an \(m\)-dimensional vector of parameters. In a GAMLSS framework \citet{Rigby2005} the elements \(j=1,...,m\) of \(\bm \theta_t\) are modelled via
\begin{equation} \label{eq:basicGAM}
    g_j(\theta_{j,t})=\mathbf{A}_{j,t} \bm{\beta}_j + \sum_{i} f_{j,i}({\bm x}^{S_{j,i}}_t), \;\;\; \text{for} \;\;\; j = 1, \dots, m,
\end{equation}
where \(g_j\) is a monotonic function, \(\mathbf{A}_{j,t}\) is the \(t\)-th row of the design matrix \(\mathbf{A}_j\), \(\bm \beta_j\) is a vector of regression coefficients, \(\bm x_t\) is a \(d\)-dimensional vector of covariates and \(S_{j,i} \subset \{1, \dots, d\}\). If \(S_{j,i} = \{1, 3\}\), then following our notation \({\bm x}_{t}^{S_{j,i}}\) is a two dimensional vector formed by the first and third element of \(\bm x_t\). Each \(f_{j,i}\) is a smooth function, constructed as
\begin{equation}
    f_{j,i}(\bm x^{S_{j,i}}) = \sum_{k=1}^{K_{j,i}} b^{ji}_k (\bm x^{S_{j,i}}) \beta_k^{ji},
\end{equation}
where \(b^{ji}_k\) are spline basis functions of dimension \(\vert S_{j,i} \vert\), while \(\beta_k^{ji}\) are regression coefficients. The smoothness of each \(f_{j,i}\) is controlled via ridge penalties, the definition of smoothness being dependent on the type of effect and penalty being used. See \citet{Wood2017} for a detailed introduction to GAM/GAMLSS models, smoothing splines bases and penalties.
